Computer-supported Analysis of Modern Variants of G\"odel's Ontological Argument

### 2014-10-18 Papers/InProgress/LogicAndReligionCongress/Abstract.tex
minor modifications

%\section{The Modal Collapse}
The axioms in G\"odel's ontological argument
\cite{GoedelNotes,ScottNotes} imply what is called the modal collapse
\cite{Sobel1987,SobelBook} (formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. God(x)$ as intended). This fact, which has
recently been confirmed with higher-order automated theorem provers
\cite{J30,ECAI-2014}, has led to strong criticism of the argument, and
it has therefore stimulated attempts to remedy the problem. Some
authors, including Hajek
\cite{Hajek_der_Mathematiker,Hajek_Magari_and_others}, suggest the use
of cautious instead of full comprehension principles in order to
escape the modal collapse. Others, including Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordahl \cite{}, have proposed to adapt G\"odel's axioms (in
particular, controversial axiom A1) and definitions.

We have studied --- with computer-support --- the emendations proposed
by Anderson and Bjordahl.  More precisely, we have formalized their
emendations within the proof assistant Isabelle/HOL \cite{Isabelle},
and we have then automated (different variations of) their emendations
with the higher-order reasoners Leo-II \cite{}, Satallax \cite{},
Metis \cite{} and Nitpick \cite{}; these systems are available in
Isabelle/HOL.  Our formalizations\footnote{The formalizations are
  available in the subdirectories \url{Anderson} and \url{Bjordahl} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
employ the embedding of higher-order modal logic (HOML) in classical
higher-order logic (HOL) as introduced and employed in our previous
work \cite{J30,ECAI-2014}. In this approach full comprehension is
naturally ``built-in'' since the underlying HOL supports
$\lambda$-abstraction.

We summarize the main results of our experiments, and we start with
Anderson's emendation (cf. Appendix \ref{apx:Anderson}), which we have
analysed for different domain conditions. These variations were
motivated by various comments on Anderson's work in the literature. 
\begin{itemize}
\item Constant domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_constant_domain.thy}} (the individual
  domains remain the same in all worlds): G\"odel's theorems C1, and
  T3 can easily be automated (in logics K and KB, respectively) ,
  G\"odel's axioms A4 and A5 are proven redundant (the former one in
  logic K4B, the latter already in K); a trivial countermodel (two
  worlds, one individual; this is simpler than
          the countermodel with two individuals and two worlds
          presented by Anderson, 1990) to MC generated by Nitpick (for all mentioned
  logics); this model also proves consistency.
\item Varying domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_varying_domain.thy}} (the domains for
  indivduals may vary from world to world, all other domains remain
  constant; different authors have pointed out that varying domain
  quantifiers should be used in Anderson's work; here we consistently
  used varying domain for all individual quantifiers; the varying
  domain quantifiers have been modeled by using an explicit
  ``Existence'' relation between indiviudals and worlds): we obtain
  the same results as  above
\item Mixed variant\footnote{Cf. Isabelle/HOL source files
    \url{Anderson/Anderson_mixed_domain_1/2.thy}} (varying domain
  quantifiers are used exclusively in the definitions of essence and
  NE; see the comments by Fitting): G\"odel's axioms A4 and A5 remain
  redundant (the former one in
  logic K4B, the latter already in K). The situation changes when only
  NE is modeled with a varying domain quantifier. Now, Nitpick reports
  a countermodel. 
\end{itemize}

@@

% Various slightly different versions of
% axioms and definitions have been considered by G\"{o}del and by
% several philosophers who commented on his proof
% (cf.~\cite{sobel2004logic,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).

@@

\begin{thebibliography}{1}

\bibitem{Adams}
R.M. Adams, `Introductory note to *1970', in {\em {Kurt G\"odel: Collected
  Works Vol. 3: Unpubl. Essays and Letters}}, Oxford Univ. Press, (1995).

\bibitem{AndersonGettings}
A.C. Anderson and M.~Gettings, `G\"odel ontological proof revisited', in {\em
  {G\"odel'96: Logical Foundations of Mathematics, Computer Science, and
  Physics: Lecture Notes in Logic 6}},  167--172, {Springer}, (1996).

\bibitem{anderson90:_some_emend_of_goedel_ontol_proof}
C.A. Anderson, `Some emendations of {G{\"o}del's} ontological proof', {\em
  Faith and Philosophy}, {\bf 7}(3), (1990).

\bibitem{Andrews:gmae72}
P.B. Andrews, `General models and extensionality', {\em Journal of Symbolic
  Logic}, {\bf 37}(2),  395--397, (1972).

\bibitem{andrewsSEP}
P.B. Andrews, `Church's type theory', in {\em The Stanford Encyclopedia of
  Philosophy}, ed., E.N. Zalta, spring 2014 edn., (2014).

\bibitem{C36}
C.~Benzm{\"u}ller, `{HOL} based universal reasoning', in {\em Handbook of the
  4th World Congress and School on Universal Logic}, ed., J.Y. Beziau~et al.,
  pp. 232--233, Rio de Janeiro, Brazil, (2013).

\bibitem{B5}
C.~Benzm{\"u}ller and D.~Miller, `Automation of higher-order logic', in {\em
  Handbook of the History of Logic, Volume 9 --- Logic and Computation},
  Elsevier, (2014).
\newblock Forthcoming; preliminary version available at
  {http://christoph-benzmueller.de/papers/B5.pdf}.

\bibitem{C34}
C.~Benzm{\"u}ller, J.~Otten, and Th. Raths, `Implementing and evaluating
  provers for first-order modal logics', in {\em Proc. of the 20th European
  Conference on Artificial Intelligence (ECAI)}, pp. 163--168, (2012).

\bibitem{B9}
C.~Benzm{\"u}ller and L.C. Paulson, `Exploring properties of normal multimodal
  logics in simple type theory with {LEO-II}', in {\em {Festschrift in Honor of
  {Peter B. Andrews} on His 70th Birthday}}, ed., C.~Benzm{\"u}ller~et al.,
  386--406, College Publications, (2008).

\bibitem{J23}
C.~Benzm{\"u}ller and L.C. Paulson, `Quantified multimodal logics in simple
  type theory', {\em Logica Universalis}, {\bf 7}(1),  7--20, (2013).

\bibitem{LEO-II}
C.~Benzm{\"u}ller, F.~Theiss, L.~Paulson, and A.~Fietzke, `{LEO-II} - a
  cooperative automatic theorem prover for higher-order logic', in {\em
  Proc.~of IJCAR 2008}, number 5195 in LNAI, pp. 162--170. Springer, (2008).

\bibitem{J30}
C.~Benzm{\"u}ller and B.~Woltzenlogel-Paleo, `Formalization, mechanization and
  automation of {G{\"o}del's} proof of {God's} existence', {\em
  arXiv:1308.4526}, (2013).

\bibitem{J28}
C.~Benzm\"uller and B.~Woltzenlogel-Paleo, `{G{\"o}del's God in Isabelle/HOL}',
  {\em Archive of Formal Proofs}, (2013).

\bibitem{W50}
C.~Benzm\"uller and B.~Woltzenlogel-Paleo, `G\"odel's {God} on the computer',
  in {\em Proceedings of the 10th International Workshop on the Implementation
  of Logics}, EPiC Series. EasyChair, (2013).
\newblock Invited abstract.

\bibitem{Coq}
Y.~Bertot and P.~Casteran, {\em {Interactive Theorem Proving and Program
  Development}}, Springer, 2004.

\bibitem{Nitpick}
J.C. Blanchette and T.~Nipkow, `Nitpick: A counterexample generator for
  higher-order logic based on a relational model finder', in {\em Proc. of ITP
  2010}, number 6172 in LNCS, pp. 131--146. Springer, (2010).

\bibitem{Satallax}
C.E. Brown, `Satallax: An automated higher-order prover', in {\em Proc. of
  IJCAR 2012}, number 7364 in LNAI, pp. 111 -- 117. Springer, (2012).

\bibitem{ContemporaryBibliography}
R.~Corazzon.
\newblock Contemporary~bibliography~on~ontological~arguments: {\scriptsize
  \url{http://www.ontology.co/biblio/ontological-proof-contemporary-biblio.htm}}.

\bibitem{Fitting}
M.~Fitting, {\em Types, Tableaux and G\"odel's God}, Kluwer, 2002.

\bibitem{fitting98}
M.~Fitting and R.L. Mendelsohn, {\em First-Order Modal Logic}, volume 277 of
  {\em Synthese Library}, Kluwer, 1998.

\bibitem{Gallin75}
D.~Gallin, {\em Intensional and Higher-Order Modal Logic}, North-Holland, 1975.

\bibitem{garbacz12:_prover_simpl_expal_away}
P.~Garbacz, `{PROVER9's} simplifications explained away', {\em Australasian
  Journal of Philosophy}, {\bf 90}(3),  585--592, (2012).

\bibitem{GoedelNotes}
K.~G\"odel, {\em Appx.A: Notes in Kurt G\"odel's Hand},  144--145.
\newblock In  \cite{sobel2004logic}, 2004.

\bibitem{Henkin50}
L.~Henkin, `Completeness in the theory of types', {\em Journal of Symbolic
  Logic}, {\bf 15}(2),  81--91, (1950).

\bibitem{homl}
R.~Muskens, `{Higher Order Modal Logic}', in {\em Handbook of Modal Logic},
  ed., P~Blackburn~et al.,  621--653, Elsevier, Dordrecht, (2006).

\bibitem{Isabelle}
T.~Nipkow, L.C. Paulson, and M.~Wenzel, {\em {Isabelle/HOL: A Proof Assistant
  for Higher-Order Logic}}, number 2283 in LNCS, Springer, 2002.

\bibitem{oppenheimera11}
P.E. Oppenheimer and E.N. Zalta, `A computationally-discovered simplification
  of the ontological argument', {\em Australasian Journal of Philosophy}, {\bf
  89}(2),  333--349, (2011).

\bibitem{rushby13}
J.~Rushby, `The ontological argument in {PVS}', in {\em Proc.~of CAV Workshop
  ``Fun With Formal Methods''}, St. Petersburg, Russia,, (2013).

\bibitem{Schulz:AICOM-2002}
S.~Schulz, `E -- a brainiac theorem prover', {\em {AI Communications}}, {\bf
  15}(2),  111--126, (2002).

\bibitem{ScottNotes}
D.~Scott, {\em Appx.B: Notes in Dana Scott's Hand},  145--146.
\newblock In  \cite{sobel2004logic}, 2004.

\bibitem{sobel2004logic}
J.H. Sobel, {\em Logic and Theism: Arguments for and Against Beliefs in God},
  Cambridge U. Press, 2004.

\bibitem{sutcliffe2009tptp}
G.~Sutcliffe, `The {TPTP} problem library and associated infrastructure', {\em
  Journal of Automated Reasoning}, {\bf 43}(4),  337--362, (2009).

\bibitem{J22}
G.~Sutcliffe and C.~Benzm{\"u}ller, `Automated reasoning in higher-order logic
  using the {TPTP THF} infrastructure.', {\em Journal of Formalized Reasoning},
  {\bf 3}(1),  1--27, (2010).

\bibitem{J29}
B.~Woltzenlogel-Paleo and C.~Benzm{\"u}ller, `Automated verification and
  reconstruction of {G\"odel's} proof of {God's} existence', {\em OCG J.},
  (2013).

\end{thebibliography}

>>>>

%\section{The Modal Collapse}
The axioms in G\"odel's ontological argument
\cite{GoedelNotes,ScottNotes} imply what is called the modal collapse
\cite{Sobel1987,SobelBook2004} (formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. God(x)$ as intended). This fact, which has
recently been confirmed with higher-order automated theorem provers
\cite{J30,ECAI-2014}, has led to strong criticism of the argument, and
it has therefore stimulated attempts to remedy the problem. Some
authors, including Hajek
\cite{Hajek_der_Mathematiker,Hajek_Magari_and_others}, suggest the use
of cautious instead of full comprehension principles in order to
escape the modal collapse. Others, including Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordahl \cite{Bjordahl}, have proposed to adapt G\"odel's axioms (in
particular, controversial axiom A1) and definitions.

We have studied --- with computer-support --- the emendations proposed
by Anderson and Bjordahl.  More precisely, we have formalized their
emendations within the proof assistant Isabelle/HOL \cite{Isabelle},
and we have then automated (different variations of) their emendations
with the higher-order reasoners Leo-II \cite{C26}, Satallax
\cite{brown2012satallax}, Metis \cite{Hurd03first-orderproof} and
Nitpick \cite{BN10}; these systems are available in Isabelle/HOL.  Our
formalizations\footnote{The formalizations are available in the
  subdirectories \url{Anderson} and \url{Bjordahl} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
employ the embedding of higher-order modal logic (HOML) in classical
higher-order logic (HOL) as introduced and employed in our previous
work \cite{J30,ECAI-2014}. In this approach full comprehension is
naturally ``built-in'' since the underlying HOL supports
$\lambda$-abstraction. The assumption of full comprehension, up to our
understanding, is
in-line with Anderson and Bjordahl (?).

We summarize the main results of our experiments, and we start with
Anderson's emendation (cf. Appendix \ref{apx:Anderson}), which we have
analysed for different domain conditions. These variations were
motivated by various comments on Anderson's work in the literature
(maybe we should be more explicit).
\begin{itemize}
\item Constant domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_constant_domain.thy}} (the individual
  domains remain the same in all worlds): G\"odel's theorems C1, and
  T3 can easily be automated (in logics K and KB, respectively) ,
  G\"odel's axioms A4 and A5 are proven redundant (the former one in
  logic K4B, the latter already in K); a trivial countermodel (two
  worlds, one individual; this is simpler than
          the countermodel with two individuals and two worlds
          presented by Anderson, 1990) to MC generated by Nitpick (for all mentioned
  logics); this model also proves consistency.
\item Varying domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_varying_domain.thy}} (the domains for
  indivduals may vary from world to world, all other domains remain
  constant; different authors have pointed out that varying domain
  quantifiers should be used in Anderson's work; here we consistently
  used varying domain for all individual quantifiers; the varying
  domain quantifiers have been modeled by using an explicit
  ``Existence'' relation between indiviudals and worlds): we obtain
  the same results as  above
\item Mixed variant\footnote{Cf. Isabelle/HOL source files
    \url{Anderson/Anderson_mixed_domain_1/2.thy}} (varying domain
  quantifiers are used exclusively in the definitions of essence and
  NE; see the comments by Fitting): G\"odel's axioms A4 and A5 remain
  redundant (the former one in
  logic K4B, the latter already in K). The situation changes when only
  NE is modeled with a varying domain quantifier. Now, Nitpick reports
  a countermodel. 
\end{itemize}

Need to do Bjordahl \ldots.

@@

% Various slightly different versions of
% axioms and definitions have been considered by G\"{o}del and by
% several philosophers who commented on his proof
% (cf{sobel2004logic,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).

@@

\bibliographystyle{plain}
\bibliography{Bibliography}

% \begin{thebibliography}{1}

% \bibitem{Adams}
% R.M. Adams, `Introductory note to *1970', in {\em {Kurt G\"odel: Collected
%   Works Vol. 3: Unpubl. Essays and Letters}}, Oxford Univ. Press, (1995).

% \bibitem{AndersonGettings}
% A.C. Anderson and M.~Gettings, `G\"odel ontological proof revisited', in {\em
%   {G\"odel'96: Logical Foundations of Mathematics, Computer Science, and
%   Physics: Lecture Notes in Logic 6}},  167--172, {Springer}, (1996).

% \bibitem{anderson90:_some_emend_of_goedel_ontol_proof}
% C.A. Anderson, `Some emendations of {G{\"o}del's} ontological proof', {\em
%   Faith and Philosophy}, {\bf 7}(3), (1990).

% \bibitem{Andrews:gmae72}
% P.B. Andrews, `General models and extensionality', {\em Journal of Symbolic
%   Logic}, {\bf 37}(2),  395--397, (1972).

% \bibitem{andrewsSEP}
% P.B. Andrews, `Church's type theory', in {\em The Stanford Encyclopedia of
%   Philosophy}, ed., E.N. Zalta, spring 2014 edn., (2014).

% \bibitem{C36}
% C.~Benzm{\"u}ller, `{HOL} based universal reasoning', in {\em Handbook of the
%   4th World Congress and School on Universal Logic}, ed., J.Y. Beziau~et al.,
%   pp. 232--233, Rio de Janeiro, Brazil, (2013).

% \bibitem{B5}
% C.~Benzm{\"u}ller and D.~Miller, `Automation of higher-order logic', in {\em
%   Handbook of the History of Logic, Volume 9 --- Logic and Computation},
%   Elsevier, (2014).
% \newblock Forthcoming; preliminary version available at
%   {http://christoph-benzmueller.de/papers/B5.pdf}.

% \bibitem{C34}
% C.~Benzm{\"u}ller, J.~Otten, and Th. Raths, `Implementing and evaluating
%   provers for first-order modal logics', in {\em Proc. of the 20th European
%   Conference on Artificial Intelligence (ECAI)}, pp. 163--168, (2012).

% \bibitem{B9}
% C.~Benzm{\"u}ller and L.C. Paulson, `Exploring properties of normal multimodal
%   logics in simple type theory with {LEO-II}', in {\em {Festschrift in Honor of
%   {Peter B. Andrews} on His 70th Birthday}}, ed., C.~Benzm{\"u}ller~et al.,
%   386--406, College Publications, (2008).

% \bibitem{J23}
% C.~Benzm{\"u}ller and L.C. Paulson, `Quantified multimodal logics in simple
%   type theory', {\em Logica Universalis}, {\bf 7}(1),  7--20, (2013).

% \bibitem{LEO-II}
% C.~Benzm{\"u}ller, F.~Theiss, L.~Paulson, and A.~Fietzke, `{LEO-II} - a
%   cooperative automatic theorem prover for higher-order logic', in {\em
%   Proc.~of IJCAR 2008}, number 5195 in LNAI, pp. 162--170. Springer, (2008).

% \bibitem{J30}
% C.~Benzm{\"u}ller and B.~Woltzenlogel-Paleo, `Formalization, mechanization and
%   automation of {G{\"o}del's} proof of {God's} existence', {\em
%   arXiv:1308.4526}, (2013).

% \bibitem{J28}
% C.~Benzm\"uller and B.~Woltzenlogel-Paleo, `{G{\"o}del's God in Isabelle/HOL}',
%   {\em Archive of Formal Proofs}, (2013).

% \bibitem{W50}
% C.~Benzm\"uller and B.~Woltzenlogel-Paleo, `G\"odel's {God} on the computer',
%   in {\em Proceedings of the 10th International Workshop on the Implementation
%   of Logics}, EPiC Series. EasyChair, (2013).
% \newblock Invited abstract.

% \bibitem{Coq}
% Y.~Bertot and P.~Casteran, {\em {Interactive Theorem Proving and Program
%   Development}}, Springer, 2004.

% \bibitem{Nitpick}
% J.C. Blanchette and T.~Nipkow, `Nitpick: A counterexample generator for
%   higher-order logic based on a relational model finder', in {\em Proc. of ITP
%   2010}, number 6172 in LNCS, pp. 131--146. Springer, (2010).

% \bibitem{Satallax}
% C.E. Brown, `Satallax: An automated higher-order prover', in {\em Proc. of
%   IJCAR 2012}, number 7364 in LNAI, pp. 111 -- 117. Springer, (2012).

% \bibitem{ContemporaryBibliography}
% R.~Corazzon.
% \newblock Contemporary~bibliography~on~ontological~arguments: {\scriptsize
%   \url{http://www.ontology.co/biblio/ontological-proof-contemporary-biblio.htm}}.

% \bibitem{Fitting}
% M.~Fitting, {\em Types, Tableaux and G\"odel's God}, Kluwer, 2002.

% \bibitem{fitting98}
% M.~Fitting and R.L. Mendelsohn, {\em First-Order Modal Logic}, volume 277 of
%   {\em Synthese Library}, Kluwer, 1998.

% \bibitem{Gallin75}
% D.~Gallin, {\em Intensional and Higher-Order Modal Logic}, North-Holland, 1975.

% \bibitem{garbacz12:_prover_simpl_expal_away}
% P.~Garbacz, `{PROVER9's} simplifications explained away', {\em Australasian
%   Journal of Philosophy}, {\bf 90}(3),  585--592, (2012).

% \bibitem{GoedelNotes}
% K.~G\"odel, {\em Appx.A: Notes in Kurt G\"odel's Hand},  144--145.
% \newblock In  \cite{sobel2004logic}, 2004.

% \bibitem{Henkin50}
% L.~Henkin, `Completeness in the theory of types', {\em Journal of Symbolic
%   Logic}, {\bf 15}(2),  81--91, (1950).

% \bibitem{homl}
% R.~Muskens, `{Higher Order Modal Logic}', in {\em Handbook of Modal Logic},
%   ed., P~Blackburn~et al.,  621--653, Elsevier, Dordrecht, (2006).

% \bibitem{Isabelle}
% T.~Nipkow, L.C. Paulson, and M.~Wenzel, {\em {Isabelle/HOL: A Proof Assistant
%   for Higher-Order Logic}}, number 2283 in LNCS, Springer, 2002.

% \bibitem{oppenheimera11}
% P.E. Oppenheimer and E.N. Zalta, `A computationally-discovered simplification
%   of the ontological argument', {\em Australasian Journal of Philosophy}, {\bf
%   89}(2),  333--349, (2011).

% \bibitem{rushby13}
% J.~Rushby, `The ontological argument in {PVS}', in {\em Proc.~of CAV Workshop
%   ``Fun With Formal Methods''}, St. Petersburg, Russia,, (2013).

% \bibitem{Schulz:AICOM-2002}
% S.~Schulz, `E -- a brainiac theorem prover', {\em {AI Communications}}, {\bf
%   15}(2),  111--126, (2002).

% \bibitem{ScottNotes}
% D.~Scott, {\em Appx.B: Notes in Dana Scott's Hand},  145--146.
% \newblock In  \cite{sobel2004logic}, 2004.

% \bibitem{sobel2004logic}
% J.H. Sobel, {\em Logic and Theism: Arguments for and Against Beliefs in God},
%   Cambridge U. Press, 2004.

% \bibitem{sutcliffe2009tptp}
% G.~Sutcliffe, `The {TPTP} problem library and associated infrastructure', {\em
%   Journal of Automated Reasoning}, {\bf 43}(4),  337--362, (2009).

% \bibitem{J22}
% G.~Sutcliffe and C.~Benzm{\"u}ller, `Automated reasoning in higher-order logic
%   using the {TPTP THF} infrastructure.', {\em Journal of Formalized Reasoning},
%   {\bf 3}(1),  1--27, (2010).

% \bibitem{J29}
% B.~Woltzenlogel-Paleo and C.~Benzm{\"u}ller, `Automated verification and
%   reconstruction of {G\"odel's} proof of {God's} existence', {\em OCG J.},
%   (2013).

%\end{thebibliography}
